---
title: "Session 4: Homework 2"
author: "Group 26"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---




```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
library(infer)
```



# Climate change and temperature anomalies 


```{r weather_data, cache=TRUE, include=FALSE}

weather <- 
  read_csv("https://data.giss.nasa.gov/gistemp/tabledata_v3/NH.Ts+dSST.csv", 
           skip = 1, 
           na = "***")

```

Beforehand, we downloaded a data set on weather anomalies in the past. We will make some modifications to it and analyze it.
Let's start by selecting the months from January to December and drop the "periods" variables. 

```{r tidyweather}
tidyweather<- weather %>% 
  select(Year,Jan:Dec) %>% 
  pivot_longer(2:13, names_to="month", 
               values_to = "delta")
  
```

Our tidyweather data frame has now three variables now, one each for 

1. year, 
1. month, and 
1. delta, or temperature deviation.

## Plotting Information

Let us plot the data using a time-series scatter plot, and add a trendline. To do that, we first need to create a new variable called `date` in order to ensure that the `delta` values are plot chronologically. 


```{r scatter_plot}

tidyweather <- tidyweather %>%
  mutate(date = ymd(paste(as.character(Year), month, "1")),
         month = month(date, label=TRUE),
         year = year(date))

ggplot(tidyweather, aes(x=date, y = delta))+
  geom_point(alpha=0.8)+
  geom_smooth(color="red") +
  theme_bw() +
  labs (title = "Weather anomalies turning positive and increasing in the last 50 years",
    subtitle = "Weather Anomalies", 
    x = "Year", 
    y= "Temperature Delta"
  )

```




It is sometimes useful to group data into different time periods to study historical data. For example, we often refer to decades such as 1970s, 1980s, 1990s etc. to refer to a period of time. NASA calculates a temperature anomaly, as difference form the base period of 1951-1980. The code below creates a new data frame called `comparison` that groups data in five time periods: 1881-1920, 1921-1950, 1951-1980, 1981-2010 and 2011-present. 

We remove data before 1800 and before using `filter`. Then, we use the `mutate` function to create a new variable `interval` which contains information on which period each observation belongs to. We can assign the different periods using `case_when()`.


```{r intervals, eval=TRUE}

comparison <- tidyweather %>% 
  filter(Year>= 1881) %>%     #remove years prior to 1881
  #create new variable 'interval', and assign values based on criteria below:
  mutate(interval = case_when(
    Year %in% c(1881:1920) ~ "1881-1920",
    Year %in% c(1921:1950) ~ "1921-1950",
    Year %in% c(1951:1980) ~ "1951-1980",
    Year %in% c(1981:2010) ~ "1981-2010",
    TRUE ~ "2011-present"
  ))

```


Now that we have the `interval` variable, we can create a density plot to study the distribution of monthly deviations (`delta`), grouped by the different time periods we are interested in. 

```{r density_plot, eval=TRUE}

ggplot(comparison, aes(x=delta, fill=interval))+
  geom_density(alpha=0.2) +   #density plot with tranparency set to 20%
  theme_bw() +                #theme
  labs (
    title = "More recent periods see increasingly positive weather anomalies",
     subtitle = "Distribution of weather anomalies by time period",
    y     = "Density"         #changing y-axis label to sentence case
    ,x = "Temperature Delta"
  )

```


